---
title: "Real data applications"
output: html_notebook
---



# Startup
Load libraries

```{r}
library(ggplot2)
library(Matrix)
library(gridExtra)
```



# Load and format data
Conversion to RDS - pbmc  
  
In folder 10x_pbmc, run convert.sh

Conversion to RDS - Ye dataset1  
  
In folder Ye_doublet_dataset1, run convert.sh

Conversion to RDS - Ye dataset2

In folder Ye_doublet_dataset2, run convert.sh



Load RDS'

```{r}
pbmc = readRDS("../realdata/10x_pbmc/10x_pbmc.mat.RDS")
ye_ds1 = readRDS("../realdata/Ye_doublet_dataset1/Ye_doublet_dataset1.mat.RDS")
ye_ds2 = readRDS("../realdata/Ye_doublet_dataset2/Ye_doublet_dataset2.mat.RDS")

ds_list = list("PBMC" = pbmc, "Ye1"= ye_ds1, "Ye2" = ye_ds2)
```


# Run and load empty droplet modeling
Fitting model for empty droplets (run in terminal, RStudio might crash)

Rscript ../betterfit/gamma_fitter.R ./10x_pbmc/10x_pbmc.mat.RDS 10x_pbmc_res
Rscript ../betterfit/gamma_fitter.R ./Ye_doublet_dataset1/Ye_doublet_dataset1.
mat.RDS Ye_doublet_dataset1
Rscript ../betterfit/gamma_fitter.R ./Ye_doublet_dataset2/Ye_doublet_dataset2.
mat.RDS Ye_doublet_dataset2



Load modeling results

```{r}
pbmc_res = read.table("../realdata/10x_pbmc/10x_pbmc_res.tsv", 
                       sep = "\t", header = T, row.names = 1)
ye1_res = read.table("../realdata/Ye_doublet_dataset1/Ye_doublet_dataset1.tsv", 
                       sep = "\t", header = T, row.names = 1)
ye2_res = read.table("../realdata/Ye_doublet_dataset2/Ye_doublet_dataset2.tsv", 
                       sep = "\t", header = T, row.names = 1)
fdr_list = list("PBMC" = pbmc_res, "Ye1" = ye1_res, "Ye2" = ye2_res)
```



Plot FDR distributions

```{r}
pdf("FDR_distributions.pdf", useDingbats = F, height = 5, width = 5)
for(n in names(fdr_list)){
  plot_df = fdr_list[[n]]
  dist_plt = ggplot(plot_df, aes(x = FDR))+
    geom_density()+
    scale_x_continuous(expand = c(0,0))+
    scale_y_continuous(expand = c(0,0))+
    geom_vline(xintercept = 0.05, colour = "red")+
    ggtitle(n)+
    theme_classic()+
    theme(axis.text = element_text(colour = "black"))
  print(dist_plt)
}
dev.off()
```


Windows of total UMIs, empty/not-empty ratios

```{r}
pdf("./UMI_windows_ratio_sig.pdf", useDingbats = F, height = 5, width = 5)
window_list = list()
for(n in names(ds_list)){
  total.umis = colSums(ds_list[[n]])
  sorted_fdr = fdr_list[[n]][colnames(ds_list[[n]]),]
  isSig = sorted_fdr<0.05
  
  windows = Hmisc::cut2(total.umis, g = 200)
  ratios = unlist(lapply(tapply(factor(isSig, levels = c("TRUE", "FALSE")), 
                                windows, table), 
                         function(x) x[1]/x[2]))
  
  plot_df = data.frame("windows" = levels(windows),
                       "ratios" = ratios)
  dist_plt = ggplot(plot_df, aes(x = windows, y = ratios, group = 1))+
    geom_point()+geom_line()+
    ggtitle(n)+
    theme_classic()+
    theme(axis.text = element_text(colour = "black"))
  
  print(dist_plt)
  window_list[[n]] = dist_plt
}
dev.off()
lapply(window_list , grid.arrange)
```



# Dimensionality reduction
Run PCA on all datasets. Using top 5000 CV genes

```{r}
pca_list = list()
sub_obs_list = list()
for(n in names(ds_list)){
  for(umitot in c(100, 250, 1000, 10000000)){
      sub_obs = ds_list[[n]][,!is.na(fdr_list[[n]]$FDR)]
      sub_obs = sub_obs[,fdr_list[[n]]$FDR[!is.na(fdr_list[[n]]$FDR)]<0.05 |
                               colSums(sub_obs)>umitot]
      sd_genes = apply(sub_obs, 1, sd)
      cv_genes = sd_genes/rowSums(sub_obs)
      cv_genes = cv_genes[order(cv_genes, decreasing = T)]
      genes_var = names(sd_genes)[1:5000]
      sub_obs = sub_obs[genes_var,]
      sub_obs_list[[paste0(n, "_", umitot)]] = sub_obs
      
      scaled_data = t(scale(t(sub_obs)/colSums(sub_obs), center = T, scale = F))
      pca_list[[paste0(n, "_", umitot)]] = irlba::prcomp_irlba(t(scaled_data), 
                                                               n = 50)
  }
}
saveRDS(pca_list, file = "pca_list.RDS")
```



Plot PCA

```{r}
pdf("./PCA_plots.pdf", height = 5, width = 5, useDingbats = F)
pca_plt_list = list()
for(n in names(pca_list)){
  nds = strsplit(n, "_")[[1]][1]
  plot_df = data.frame(pca_list[[n]]$x)
  plot_df$empty = ifelse(colnames(sub_obs_list[[n]]) %in% rownames(fdr_list[[nds]])[fdr_list[[nds]]$FDR>=0.05], 
                         "empty", "not-empty")
  
  #plot_df = plot_df[order(plot_df$empty, decreasing = F),]
  
  pca_plt = ggplot(plot_df, aes(x = PC1, y = PC2, colour = empty, shape = empty))+
    geom_point()+
    ggtitle(n)+
    scale_shape_manual(values = c(1, 19))+
    scale_colour_manual(values = c("black", "firebrick"))+
    theme_classic()+
    theme(legend.position = "bottom",
          axis.text = element_text(colour = "black"))
  print(pca_plt)
  write.table(pca_list[[n]]$x, file = paste0("./pca_results", n, ".txt"), 
              sep = ",", row.names = F, col.names = T, quote = F)
  
  pca_plt_list[[n]] = pca_plt
}
dev.off()
lapply(pca_plt_list, grid.arrange)
```



Run tSNE (in python because it is faster)

```{python}
from sklearn.manifold import TSNE
import pandas as pd

from os import listdir
from os.path import isfile, join
mypath = "./"
onlyfiles = [f for f in listdir(mypath) if isfile(join(mypath, f)) and "pca" in f]

print onlyfiles

for f in onlyfiles:
  pca_res = pd.read_csv(f, index_col=0, header = 0)
  
  model = TSNE(n_components=2, random_state=0)
  
  tsne_res = model.fit_transform(pca_res) 
  
  pd.DataFrame(tsne_res).to_csv(f+"_tsne.txt")
```



Load tSNE results

```{r}
tsne_list = list()
for(n in names(pca_list)){
  tsne_list[[n]] = read.table(paste0("./pca_results", n, ".txt_tsne.txt"), 
                              sep = ",", header = T, row.names = 1)
  colnames(tsne_list[[n]]) = c("dim1", "dim2")
}
```



Plot tSNE

```{r}
pdf("./tSNE_plots.pdf", height = 5, width = 5, useDingbats = F)
tsne_plt_list = list()
for(n in names(pca_list)){
  nds = strsplit(n, "_")[[1]][1]
  plot_df = data.frame(tsne_list[[n]])
  plot_df$empty = ifelse(colnames(sub_obs_list[[n]]) %in% rownames(fdr_list[[nds]])[fdr_list[[nds]]$FDR>=0.05], 
                         "empty", "not-empty")
  
  #plot_df = plot_df[order(plot_df$empty, decreasing = F),]
  
  tsne_plt = ggplot(plot_df, aes(x = dim1, y = dim2, colour = empty, shape = empty))+
    geom_point(alpha = 0.65)+
    ggtitle(n)+
    scale_shape_manual(values = c(1, 19))+
    scale_colour_manual(values = c("black", "firebrick"))+
    theme_classic()+
    theme(legend.position = "bottom",
          axis.text = element_text(colour = "black"))
  print(tsne_plt)
  
  tsne_plt_list[[n]] = tsne_plt
}
dev.off()
lapply(tsne_plt_list, grid.arrange)
```

